Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2024-07-03 manuscript.Rmd
Update manuscript.Rmd

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment (see also @Barber2022; @Chen-Xia2023; @Kasper2022; @Morehouse2023; @Rousseau;2023; @West2023). The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

@@

@Bastian 2013 seek to account for this possibility by statistically controlling for participants’ moral outrage at the targets’ behaviour in their analysis.  They report that the relationship between trait-based dehumanization and punishment remains even when moral outrage is controlled for.  While this is interesting and suggestive of the independent effects of dehumanization, it cannot fully address the conceptual weaknesses in how dehumanization was operationalised.  A more convincing way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but vary from socially desirable to undesirable [@Over2021a, -@Over2021b].  Previous research conducted by @Enock2021b has established that undesirable character traits such as jealous, spiteful and bitter are considered unique to humans and socially undesirable.  Across three intergroup contexts, the researchers found that participants attributed socially desirable human traits more strongly to the ingroup and socially undesirable traits more strongly to the outgroup; see also [@Decker2023; @Enock2021a; @Enock2022, -@Enock2023].  Enock and colleagues -@Enock2021b concluded that intergroup preference may better explain apparent evidence for trait-based dehumanization.  However, it is not yet clear how the attribution of uniquely human character traits relates to harm.  Addressing this question is crucial to understanding the extent to which the dual model of dehumanization can help explain real-world discrimination and negativity.

>>>>

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment (see also @Barber2022; @Chen-Xia2023; @Kasper2022; @Morehouse2023; @Rousseau2023; @West2022). The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

@@

@Bastian2013 seek to account for this possibility by statistically controlling for participants’ moral outrage at the targets’ behaviour in their analysis.  They report that the relationship between trait-based dehumanization and punishment remains even when moral outrage is controlled for.  While this is interesting and suggestive of the independent effects of dehumanization, it cannot fully address the conceptual weaknesses in how dehumanization was operationalised.  A more convincing way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but vary from socially desirable to undesirable [@Over2021a, @Over2021b].  Previous research conducted by @Enock2021b has established that undesirable character traits such as jealous, spiteful and bitter are considered unique to humans and socially undesirable.  Across three intergroup contexts, the researchers found that participants attributed socially desirable human traits more strongly to the ingroup and socially undesirable traits more strongly to the outgroup; see also [@Decker2023; @Enock2021a; @Enock2022, @Enock2023].  Enock and colleagues [@Enock2021b] concluded that intergroup preference may better explain apparent evidence for trait-based dehumanization.  However, it is not yet clear how the attribution of uniquely human character traits relates to harm.  Addressing this question is crucial to understanding the extent to which the dual model of dehumanization can help explain real-world discrimination and negativity.
